---
title: "PA1_template.Rmd"
author: "Chris Morris"
date: "27 April 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(ggplot2)
require(dplyr)
require(timeDate)
```

## Loading and preprocessing the data
Read the data into a variable called activity using read.csv.

```{r preprocess}
activity <- read.csv("activity.csv")
activity$date <- as.Date(activity$date, "%Y-%m-%d")
head(activity)
summary(activity)
```
Note the large number of NA in the steps column

## What is mean total number of steps taken per day?
A histogram using ggplot2 of the total number of steps recorded for each day.  The warning is due to the NAs in he data which is addressed below.
```{r histogram steps per day, fig.width=10,fig.height=5}
ggplot(data = activity, aes(date, steps)) +
  stat_summary(fun.y = sum, geom = "bar") + 
  theme(axis.text.x=element_text(angle=80, hjust=1)) + 
  ggtitle("Steps per Day") + 
  xlab("Date") + 
  ylab("Total Steps") +  
  scale_y_continuous(expand = c(0, 0))
```

### Mean and median number of steps taken each day
Output a table of the mean and mediam number of steps taken per day.  Uses the summarise function from dplyr.  
```{r mean and median number of steps per day}
activity_summary <- activity %>% 
  group_by(date) %>% 
  summarise (Mean = mean(steps, na.rm = TRUE), Median = median(steps, na.rm = TRUE))
format(activity_summary)
```
Note that the median shows as zero due to the large number of 5 minute intervals with no steps

## What is the average daily activity pattern?
Output a time plot of the average number of steps taken (per interval)
```{r Time series plot of the average number of steps taken, fig.width=10,fig.height=5}
interval_summary <- activity %>% 
  group_by(interval) %>% 
  summarise (Mean = mean(steps, na.rm = TRUE))

ggplot(data = interval_summary, mapping = aes(x = interval, y = Mean)) +
  geom_line() + 
  ggtitle("Average number of steps taken (per interval)") + 
  xlab("Interval") + 
  ylab("Average Steps")
```


### The 5-minute interval that, on average, contains the maximum number of steps
```{r max interval}
max_interval <- interval_summary[which.max( interval_summary$Mean ),]
format(max_interval)
```
The interval with the highest mean is `r max_interval$interval` with an average of `r max_interval$Mean` steps

## Imputing missing values
The NA value will be replaced with the average of the interval
```{r imputing missing data}

totalna <- sum( is.na( activity$steps ) ) 

impute.mean <- function(x) {
  replace(x, is.na(x), mean(x, na.rm = TRUE))
}

activity_imputed <- activity %>%
  group_by(interval) %>%
  mutate(
    steps = impute.mean(steps)  
  )

summary(activity_imputed)
```
There were `r totalna` values replaced with the mean for that period

### Histogram of the total number of steps taken each day after missing values are imputed
```{r histogram steps per day after imputting na values, fig.width=10,fig.height=5}
ggplot(data = activity_imputed, aes(date, steps)) +
  stat_summary(fun.y = sum, geom = "bar") + 
  theme(axis.text.x=element_text(angle=80, hjust=1)) + 
  ggtitle("Steps per Day after imputting missing values") + 
  xlab("Date") + 
  ylab("Total Steps") +  
  scale_y_continuous(expand = c(0, 0))
```
### Mean and median total number of steps taken per day
```{r mean and median number of steps per day after imputting na values}
activity_imputed_summary <- activity_imputed %>% 
  group_by(date) %>% 
  summarise (Mean = mean(steps, na.rm = TRUE), Median = median(steps, na.rm = TRUE))
format(activity_imputed_summary)
``` 


## Are there differences in activity patterns between weekdays and weekends?
Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends
```{r weekend, fig.width=10,fig.height=5}
activity_imputed$weekday <- ifelse(isWeekday(activity_imputed$date, wday=1:5), "Weekday", "Weekend")

interval_imputed_summary <- activity_imputed %>% 
  group_by(interval, weekday) %>% 
  summarise (Mean = mean(steps, na.rm = TRUE))

ggplot(data = interval_imputed_summary, mapping = aes(x = interval, y = Mean)) +
  geom_line() + 
  facet_wrap(~weekday, ncol=1)+
  ggtitle("Average number of steps taken (per interval)") + 
  xlab("Interval") + 
  ylab("Average Steps")
```
While the graphs both follow similar trends, there are clear differnces between the weekdays and the weekend